Convergence rate of MAP estimates \\ for the exponential family

### 2021-10-15 aistats.tex
minor changes

\paragraph{Optimization.}
With exponential families, MAP can be seen as \emph{an instantiation of stochastic mirror descent (SMD)}, reviewed in App.~\ref{app:SMD}.
More precisely, let us re-write~\eqref{eq:defMAP} as
\alignn{
\m_n = \m_{n-1}- \lr_n (\m_{n-1} - T_n)
\label{eq:mean_update}
}
where $\lr_n := \inv{n_0 + n}$.
Now define stochastic functions $f_X(\nat) = -\log p(X \cond \nat)$ such that $\E[f_X] = f$.
If we further introduce stochastic gradients $g_n(\nat) := \nabla\logpart(\nat) - T_n = \nabla f_{X_n}(\nat)$, then~\eqref{eq:mean_update} becomes
\alignn{
	\nabla\conj(\hat \nat_{n})
	= \nabla\conj(\hat \nat_{n-1}) - \lr_n g_n(\hat \nat_{n-1}),
}
which is the update formula for SMD on $f$ with mirror map $\nabla\logpart$
and step-size schedule $\lr_n$, initialized at $\nat_0$.
In this view, MLE forgets its (arbitrary) initialization after the first step with step-size 1.
The observation MAP $\in$ SMD brings us to our second problem.
\begin{problem}[Convergence rate for SMD]
Find a convergence rate for stochastic mirror descent that applies to conjugate MAP of exponential families such as Gaussians $\cN(\meanp, \sigma^2)$.
\end{problem}

@@

A simple yet non-trivial example is the centered Gaussian distribution with unknown variance $\cN(0,\sigma^2)$.
Its log-likelihood reads $\log p(x) = -\frac{x^2}{2 \sigma^2} - \half\log(2\pi \sigma^2)$.
Defining $T(X)=X^2$ as the sufficient statistic, we get natural parameter $\nat = -\inv{2 \sigma^2} <0$, and mean parameter $\m=\E[T(X)] = \sigma^2 >0$.
Mean and natural parameters are roughly inverse of each other, i.e., $\nat = -\inv{2 \m}$.
Now we match the log-likelihood with the exponential family template to get the log-partition function, and take the conjugate to find the entropy
\aligns{
	\logpart (\nat) = - \half \log(-\nat) \quad\text{and}\quad
	\conj(\m) = - \half \log(\m)  \; ,
}
up to constants.
Both the log-partition and  the entropy are roughly negative logarithm $z\mapsto - \log(z)$.
It means the conjugate prior is the exponential family with sufficient statistic $(\nat, \log(-\nat) )$, e.g., a negative gamma distribution.
It also means $\bregman$ and $\bregmanconj$ have the same shape
\begin{align}
	\bregmanconj( \m^*; \m_n)
	&= \half \left ( \frac{\m^*}{ \m_n} - 1 - \log  \frac{\m^*}{ \m_n} \right) \; .
%	\bregman( \nat_n; \nat_* )
%	&=  \half \left ( \frac{ \nat_n}{\nat_*} - 1 - \log  \frac{ \nat_n}{\nat_*} \right) \; .
\end{align}
In Theorems~\ref{thm:varianceMLE} and~\ref{thm:varianceMAP}, we report upper bounds on the expected value of this divergence for the MLE and the MAP.
All proofs for this section are in App.~\ref{app:gaussian-variance}.
\begin{theorem}[MLE Bound]
\label{thm:varianceMLE}
	The MLE of $\cN(0,\m^*)$ is $\hat \m_n^\text{MLE} = \inv{n} \sum_i X_i^2 $.
	Its expected suboptimality is infinite when $n\leq 2$, and otherwise upper-bounded as
	\begin{align}
		 \expect{\bregmanconj( \m^*; \hat \m_n^\text{MLE}) }
			\leq \inv{2n} +\frac{2}{n(n-2)} \; .
			\label{eq:MLE_rate}
	\end{align}
\end{theorem}
This upper bound matches the asymptotic result~\eqref{eq:asymptote} that we derive in \S\ref{ssec:asymptote}.
We illustrate its numerical behavior in Figure~\ref{fig:curves}.
With the same technique, we obtain a similar bound for the multivariate generalization: the expected value is infinite whenever $n \leq d+1$ where $d$ is the dimension, and is otherwise bounded by $O(\frac{d^2}{n} + \frac{d^3}{n(n-d-1)} )$.

>>>>

\paragraph{Optimization.}
With exponential families, MAP can be linked to \emph{stochastic mirror descent (SMD)}, see App.~\ref{app:SMD}.
More precisely, let us re-write~\eqref{eq:defMAP} as
\alignn{
\m_n = \m_{n-1}- \lr_n (\m_{n-1} - T_n)
\label{eq:mean_update}
}
where $\lr_n := \inv{n_0 + n}$.
Now define stochastic functions $f_X(\nat) = -\log p(X \cond \nat)$ such that $\E[f_X] = f$.
If we further introduce stochastic gradients $g_n(\nat) := \nabla\logpart(\nat) - T_n = \nabla f_{X_n}(\nat)$, then~\eqref{eq:mean_update} becomes
\alignn{
	\nabla\conj(\hat \nat_{n})
	= \nabla\conj(\hat \nat_{n-1}) - \lr_n g_n(\hat \nat_{n-1}),
}
which is the update formula for SMD on $f$ with mirror map $\nabla\logpart$
and step-size schedule $\lr_n$, initialized at $\nat_0$.
In this view, MLE forgets its (arbitrary) initialization after the first step with step-size 1.
The observation MAP $\in$ SMD brings us to our second problem.
\begin{problem}[Convergence rate for SMD]
Find a convergence rate for stochastic mirror descent that applies to conjugate MAP of exponential families such as Gaussians $\cN(\meanp, \sigma^2)$.
\end{problem}

@@

A simple yet non-trivial example is the centered Gaussian distribution with unknown variance $\cN(0,\sigma^2)$.
Its log-likelihood reads $\log p(x) = -\frac{x^2}{2 \sigma^2} - \half\log(2\pi \sigma^2)$.
Defining $T(X)=X^2$ as the sufficient statistic, we get natural parameter $\nat = -\inv{2 \sigma^2} <0$, and mean parameter $\m=\E[T(X)] = \sigma^2 >0$.
Mean and natural parameters are roughly inverse of each other, i.e., $\nat = -\inv{2 \m}$.
Now we match the log-likelihood with the exponential family template to get the log-partition function, and take the conjugate to find the entropy
\aligns{
	\logpart (\nat) = - \half \log(-\nat) \quad\text{and}\quad
	\conj(\m) = - \half \log(\m)  \; ,
}
up to constants.
Both $A$ and  the entropy are roughly negative logarithm $z\mapsto - \log(z)$.
It means the conjugate prior is the exponential family with sufficient statistic $(\nat, \log(-\nat) )$, e.g., a negative gamma distribution.
It also means $\bregman$ and $\bregmanconj$ have the same shape
\begin{align}
	\bregmanconj( \m^*; \m_n)
	&= \half \left ( \frac{\m^*}{ \m_n} - 1 - \log  \frac{\m^*}{ \m_n} \right) \; .
%	\bregman( \nat_n; \nat_* )
%	&=  \half \left ( \frac{ \nat_n}{\nat_*} - 1 - \log  \frac{ \nat_n}{\nat_*} \right) \; .
\end{align}
In Theorems~\ref{thm:varianceMLE} and~\ref{thm:varianceMAP}, we report upper bounds on the expected value of this divergence for the MLE and the MAP.
All proofs for this section are in App.~\ref{app:gaussian-variance}.
\begin{theorem}[MLE Bound]
\label{thm:varianceMLE}
	The MLE of $\cN(0,\m^*)$ is $\hat \m_n^\text{MLE} = \inv{n} \sum_i X_i^2 $.
	Its expected suboptimality is infinite when $n\leq 2$, and otherwise upper-bounded as
	\begin{align}
		 \expect{\bregmanconj( \m^*; \hat \m_n^\text{MLE}) }
			\leq \inv{2n} +\frac{2}{n(n-2)} \; .
			\label{eq:MLE_rate}
	\end{align}
\end{theorem}
This upper bound matches the asymptotic result~\eqref{eq:asymptote} that we derive in \S\ref{ssec:asymptote}.
We illustrate its numerical behavior in Figure~\ref{fig:curves}.
With the same technique, we obtain a similar bound for the multivariate generalization: the expected value is infinite whenever $n \leq d+1$ where $d$ is the dimension, and is otherwise bounded by $O(\frac{d^2}{n} + \frac{d^3}{n(n-d-1)} )$.
